---
title: "Data Preparation - Initial Cleaning"
author: "Yunfei Yu"
date: "2026-04-11"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: yeti
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", warning = FALSE)
```

```{r libraries, message=FALSE}
library(readxl)
library(dplyr)
library(tidyverse)
library(kableExtra)
library(knitr)
library(skimr)
library(naniar)
# library(VIM)
library(gtsummary)
library(gt)
library(forcats)
library(stringr)
library(psych)
# library(polycor)
library(ggcorrplot)
library(ggplot2)

# for saving output
library(here)
```

```{r import helper functions, message=FALSE}
source("helper_functions.R")
```

# Inspect Raw Data

```{r read in raw data}
raw_data <- read_excel("Data/FINAL_ProjectOATH_DATA_2024-05-08_1639.xlsx") %>% 
  # drop redcap_event_name and redcap_survey_identifier
  select(-c(redcap_event_name, redcap_survey_identifier)) %>% 
  # drop completeness check variables
  select(-contains("complete"))

# check if record_id is unique
nrow(raw_data) == n_distinct(raw_data$record_id)

# check dimension
dim(raw_data)
```

```{r create raw data variable summary}
variable_summary <- get_missing_stats(data = raw_data, export_df = TRUE) %>% 
  mutate(category = case_when(
    str_detect(variable, "site|role|years_employed|academic") ~ "site and role",
    str_detect(variable, "age|gender|sexual_orient|sex|disability|indigenous|poc|language|race") ~ "demographics",
    variable == "record_id" ~ "identifier", 
    TRUE ~ "scale"
  ))

head(variable_summary)
```


# Label Data

## Label Site and Role

```{r label site and role data}
site_role_vars <- extract_vars(cat = "site and role")

site_role_data <- raw_data %>% 
  select(record_id, all_of(site_role_vars)) %>% 
  mutate(
    site_oath_v2 = factor(site_oath_v2,
                          levels = c(1, 2, 3, 4),
                          labels = c("Credit Valley Hospital", 
                                     "Missisauga Hospital", 
                                     "Queensway Health Centre", 
                                     "Other")
                          ),
    role_oath_v2 = factor(role_oath_v2,
                          levels = c(1, 2, 3, 4, 5, 6, 7),
                          labels = c("Administrative/Consultant",
                                     "Director",
                                     "Manager",
                                     "Professional Staff",
                                     "Senior Leadership",
                                     "Supervisor/Team Leader",
                                     "Other")
                          ),
    years_role_oath_v2 = factor(years_role_oath_v2,
                                levels = c(1, 2, 3, 4, 5),
                                labels = c("<1",
                                           "1-3",
                                           "3-6",
                                           "6-9",
                                           "10+")
                                ),
    years_employed_thp_v2 = factor(years_employed_thp_v2,
                                   levels = c(1, 2, 3, 4, 5),
                                   labels = c("<1",
                                              "1-3",
                                              "3-6",
                                              "6-9",
                                              "10+")
                                   ),
    academic_appt = factor(academic_appt,
                           levels = c(1, 0),
                           labels = c("Yes", "No"))
  )

```


## Label Demographic Variables

- **Factor**: age, disability, indigenous, poc
- **MC**: gender, sexual orientation
- **Choose all that apply**: language, race

```{r label demographics data}
demog_vars <- extract_vars(cat = "demographics")

demog_data <- raw_data %>% 
  select(record_id, all_of(demog_vars)) %>% 
  # label factor variables
  mutate(
    age_oath_v2 = factor(age_oath_v2,
                         levels = c(1, 2, 3, 4, 5, 6),
                         labels = c("<18",
                                    "18-29",
                                    "30-54",
                                    "55-64",
                                    "65+",
                                    "Prefer not to answer")
                         ),
    disability_oath_v2 = factor(disability_oath_v2,
                                levels = c(1, 2, 3),
                                labels = c("Yes", "No", "Prefer not to answer")
                                ),
    indigenous_oath_v2 = factor(indigenous_oath_v2,
                                levels = c(1, 2, 3),
                                labels = c("Yes", "No", "Prefer not to answer")
                                ),
    poc_oath_v2 = factor(poc_oath_v2,
                         levels = c(1, 2, 3),
                         labels = c("Yes", "No", "Prefer not to answer")
                         )
  ) %>% 
  # label MC variables (gender and sexual orientation): 
  # treat as missing if selected 0 or >1 categories
  mutate(gender_sum = rowSums(pick(starts_with("gender_"))) + ifelse(is.na(self_define_gender), 0, 1),
         sex_sum = rowSums(pick(starts_with("sexual_"))) + ifelse(is.na(self_define_sex1), 0, 1)) %>%
  mutate(
    gender_oath_v2 = case_when(
      gender_sum > 1 | gender_sum == 0 ~ NA,
      gender_oath_v2___1 == 1 ~ "Woman",
      gender_oath_v2___2 == 1 ~ "Man",
      gender_oath_v2___3 == 1 ~ "Trans",
      gender_oath_v2___4 == 1 ~ "Two Spirit",
      gender_oath_v2___5 == 1 ~ "Prefer to self-define",
      gender_oath_v2___6 == 1 ~ "Prefer not to answer",
      TRUE ~ NA
    ),
    sexual_orient_oath_v2 = case_when(
      sex_sum > 1 | sex_sum == 0 ~ NA,
      sexual_orient_oath_v2___1 == 1 ~ "Gay",
      sexual_orient_oath_v2___2 == 1 ~ "Lesbian",
      sexual_orient_oath_v2___3 == 1 ~ "Straight/Heterosexual",
      sexual_orient_oath_v2___4 == 1 ~ "Bisexual",
      sexual_orient_oath_v2___5 == 1 ~ "Queer",
      sexual_orient_oath_v2___6 == 1 ~ "Two Spirit",
      sexual_orient_oath_v2___7 == 1 ~ "Prefer to self-define",
      sexual_orient_oath_v2___8 == 1 ~ "Prefer not to answer",
      TRUE ~ NA
    )
  ) %>% 
  select(-starts_with(c("gender_oath_v2___", "sexual_orient_oath_v2___"))) %>% 
  select(-c(gender_sum, sex_sum)) %>% 
  # rename checkbox (choose all that apply): language
  rename(
    language_oath_v2_English = language_oath_v2___1,
    language_oath_v2_French = language_oath_v2___2,
    language_oath_v2_Indigenous = language_oath_v2___3,
    language_oath_v2_Mandarin = language_oath_v2___4,
    language_oath_v2_Cantonese = language_oath_v2___5,
    language_oath_v2_Punjabi = language_oath_v2___6,
    language_oath_v2_Hindi = language_oath_v2___7,
    language_oath_v2_Urdu = language_oath_v2___8,
    language_oath_v2_Spanish = language_oath_v2___9,
    language_oath_v2_Portuguese = language_oath_v2___10,
    language_oath_v2_Tagalog = language_oath_v2___11,
    language_oath_v2_Vietnamese = language_oath_v2___12,
    language_oath_v2_Arabic = language_oath_v2___13,
    language_oath_v2_ASL = language_oath_v2___14,
    language_oath_v2_LSQ = language_oath_v2___15,
    language_oath_v2_Other = language_oath_v2___16,
    language_oath_v2_PNA = language_oath_v2___17,
  ) %>% 
  # rename checkbox (choose all that apply): language
  rename(
    race_oath_v2_Asian = race_oath_v2___1,
    race_oath_v2_Black = race_oath_v2___2,
    race_oath_v2_LatinHispanic = race_oath_v2___3,
    race_oath_v2_MiddleEastern = race_oath_v2___4,
    race_oath_v2_Mixed = race_oath_v2___5,
    race_oath_v2_White = race_oath_v2___6,
    race_oath_v2_Other = race_oath_v2___7,
    race_oath_v2_PNA = race_oath_v2___8,
  )
```


## Check Self-Defined / Specified Values and Special Cases

- **Self-define**: gender, sexual orientation
- **Please specify**: role, language, race

```{r self defined values summary}
# joint demog and role & sites df
demog_data <- site_role_data %>% left_join(demog_data, by = "record_id")

get_category_summary(demog_data %>%
                       select(matches("spec|define")),
                     caption = "Self-defined or specified values")
```


```{r regroup self spec}
# regroup based on self-defined / specified values
# the remaining "Other" are those who didn't specify -> convert to missing

demog_final <- demog_data %>% 
  # regroup role
  mutate(role_oath_v2 = case_when(
    coalesce(str_detect(role_oath_spec, regex("educator", ignore_case = TRUE)), FALSE) ~ "Educator",
    coalesce(str_detect(role_oath_spec, regex("leader", ignore_case = TRUE)), FALSE) ~ "Supervisor/Team Leader",
    TRUE ~ role_oath_v2)
  ) %>% 
  mutate(role_oath_v2 = case_when(
    role_oath_v2 == "Other" ~ NA,
    TRUE ~ role_oath_v2)
  ) %>% 
  # language and race PNA:
  # if both PNA and any other box are checked, convert PNA to 0
  # also if PNA is checked but self-specified other values, convert PNA to 0
  mutate(language_oath_v2_PNA = case_when(
    language_oath_v2_PNA == 1 & if_any(starts_with("language_oath") & -language_oath_v2_PNA, ~ .x == 1) ~ 0,
    language_oath_v2_PNA == 1 & !is.na(language_specify_oath_v2) ~ 0,
    TRUE ~ language_oath_v2_PNA)
  ) %>% 
  mutate(race_oath_v2_PNA = case_when(
    race_oath_v2_PNA == 1 & if_any(starts_with("race_oath") & -race_oath_v2_PNA, ~ .x == 1) ~ 0,
    race_oath_v2_PNA == 1 & !is.na(race_spec) ~ 0,
    TRUE ~ race_oath_v2_PNA)
  ) %>% 
  # regroup race
  rename(race_oath_v2_BroadAsian = race_oath_v2_Asian) %>% 
  mutate(
    # combine South Asian, Indo-Caribbean and West Indian into the broader Asian / Asian heritage category
    # for aggregation purposes
    race_oath_v2_BroadAsian = case_when(
      coalesce(str_detect(race_spec, regex("Asian|Indian|Caribbean", ignore_case = TRUE)), FALSE) ~ 1,
      TRUE ~ race_oath_v2_BroadAsian
    ),
    # assuming Canadian and European are White
    race_oath_v2_White = case_when(
      coalesce(str_detect(race_spec, regex("Canadian|European", ignore_case = TRUE)), FALSE) ~ 1,
      TRUE ~ race_oath_v2_White
    )
  ) %>% 
  # language: keep only EN and FR, combine others into one category
  mutate(
    language_oath_v2_Other = case_when(
      rowSums(select(., language_oath_v2_Indigenous:language_oath_v2_Other), na.rm = TRUE) >= 1 ~ 1,
      TRUE ~ 0
    )
  ) %>% 
  select(-c(language_oath_v2_Indigenous:language_oath_v2_Arabic))

   

# create indicator for multi language & race
races <- demog_final %>% 
  select(contains("race_oath") & !matches("Mixed|Other|PNA")) %>% 
  colnames()

demog_final <- demog_final %>% 
  mutate(
    # extract the number of self specified language(s)
    temp_text_lang = str_trim(language_specify_oath_v2) %>% str_remove(",$"),
    lang_spec_count = case_when(
      is.na(temp_text_lang) | temp_text_lang == "" ~ 0,
      TRUE ~ as.numeric(lengths(str_split(temp_text_lang, ",\\s*")))
    )
  ) %>% 
  # create multilingual/racial indicator, 1 if more than one language/race, 0 otherwise
  # (not including race_spec in calculation since it is merged with other groups earlier)
  mutate(multilingual = ifelse(rowSums(pick(starts_with("language_oath"))) + lang_spec_count > 1, 
                               "Yes", "No"),
         sign_languages = case_when(
           language_oath_v2_ASL == 1 ~ "Yes",
           language_oath_v2_LSQ == 1 ~ "Yes",
           TRUE ~ "No"
         ),
         multiracial = case_when(
           rowSums(.[races]) > 1 ~ "Yes",
           race_oath_v2_Mixed == 1 ~ "Yes",
           TRUE ~ "No"
         )
  ) %>% 
  select(-matches("spec|define|temp"))
```

# Remove NAs in Scales Items

Removing participants who either

- Skipped more than 50% of the questionnaires, or

- Chose "8: Not Applicable" in more than 50% of the questions

```{r check missing in scale items df, fig.height=8}
# extract scale variables
scale_vars <- extract_vars(cat = "scale")
scale_df <- raw_data %>% 
  select(record_id, all_of(scale_vars)) %>% 
  select(-innov_lead_disc)

# calculate row-wise missing rate and proportion of 8
scale_df <- scale_df %>% 
  mutate(
    missing_rate = rowMeans(is.na(.)),
    prop_8 = rowMeans(. == 8, na.rm = TRUE)
  )

# drop observations with >50% missing rate or >50% "Not Applicable"
scale_df <- scale_df %>% 
  filter(missing_rate <= 0.5 & prop_8 <= 0.5)
# nrow(scale_df)

# check missing pattern in the remaining data
gg_miss_upset(scale_df, nsets = ncol(scale_df), mb.ratio = c(0.3, 0.7))
```

# Summary Statistics

## Demographics Summary

```{r demographic summary stats}
get_category_summary(data = demog_final, caption = "Distribution by Demographic Variables", id = "record_id")
```


## Scales 

### Distribution

Check distributions of each scale items before converting 8 to NA.

```{r explore scale distribution}
plt_scale_dist(data = scale_df, prefix = "explore_")
```


```{r exploit scale distribution}
plt_scale_dist(data = scale_df, prefix = "exploit_", num_col = 4)
```

```{r deimp scale distribution}
plt_scale_dist(data = scale_df, prefix = "deimp", num_col = 3)
```

```{r ocis scale distribution, fig.height=10}
plt_scale_dist(data = scale_df, prefix = "ocis", num_col = 6)
```

```{r ols scale distribution, fig.height=8}
plt_scale_dist(data = scale_df, prefix = "ols", num_col = 5)
```

### Scale Reliability

Note 1: treating 8 as NAs for checking reliability.

Note 2: `explore_innov_def_1`, `explore_innov_def_2`, `exploit_innov_def_1` and `exploit_innov_def_2` are added by the project team, rest of Organization Learning Ambidexterity Survey are adapted from Mom et al 2009.

```{r}
# convert 8 to missing
scale_converted <- scale_df %>% 
  mutate(across(-record_id, ~na_if(., 8)))
```


```{r check scale reliability}
explore_alpha <- scale_converted %>%
  select(starts_with("explore_")) %>% 
  psych::alpha(title = "OA Exploration")

exploit_alpha <- scale_converted %>% 
  select(starts_with("exploit")) %>%  
  psych::alpha(title = "OA Exploitation")

deimp_alpha <- scale_converted %>% 
  select(starts_with("deimp")) %>% 
  psych::alpha(title = "OA Deimplementation")

ocis_alpha <- scale_converted %>% 
  select(starts_with("ocis")) %>% 
  psych::alpha(title = "OA OCIS")

ols_alpha <- scale_converted %>% 
  select(starts_with("ols")) %>% 
  psych::alpha(title = "OA OLS")
```

```{r extract reliability resuls}
# create summary table
reliability_summary <- bind_rows(
  extract_alpha_results(explore_alpha, "Ambidexterity (Exploration)"),
  extract_alpha_results(exploit_alpha, "Ambidexterity (Exploitation)"),
  extract_alpha_results(deimp_alpha, "Deimplementation"),
  extract_alpha_results(ocis_alpha, "OCIS"),
  extract_alpha_results(ols_alpha, "OLS")
)

reliability_summary %>% 
  kable(format = "html",
        caption = "Scale Reliability") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Summary of Individual Scales

Note: treating 8 as NAs for checking reliability. "Valid n" includes non-missing and non-8 responses.

#### Exploration

```{r exploration item summary, results='asis'}
explore_item_summary <- extract_item_stats(data = scale_converted, 
                                           alpha_obj = explore_alpha, 
                                           scale_prefix = "explore", 
                                           cap = "Exploration Item Summary (Scale Raw Alpha = 0.72)")
```

#### Exploitation

```{r exploitation item summary, results='asis'}
exploit_item_summary <- extract_item_stats(data = scale_converted, 
                                           alpha_obj = exploit_alpha, 
                                           scale_prefix = "exploit", 
                                           cap = "Exploitation Item Summary (Scale Raw Alpha = 0.69)")
```


#### Deimplementation

```{r deimplementation item summary, results='asis'}
deimp_item_summary <- extract_item_stats(data = scale_converted,
                                         alpha_obj = deimp_alpha,
                                         scale_prefix = "deimp",
                                         cap = "Deimplementation Item Summary (Scale Raw Alpha = 0.88)")
```

#### OCIS

```{r ocis item summary, results='asis'}
ocis_item_summary <- extract_item_stats(data = scale_converted,
                                        alpha_obj = ocis_alpha,
                                        scale_prefix = "ocis",
                                        cap = "OCIS Item Summary (Scale Raw Alpha = 0.94)")
```


#### OLS

```{r ols item summary, results='asis'}
ols_item_summary <- extract_item_stats(data = scale_converted,
                                       alpha_obj = ols_alpha,
                                       scale_prefix = "ols",
                                       cap = "OLS Item Summary (Scale Raw Alpha = 0.92)")
```

### Polychoric Correlation

```{r polychoric correlation}
# define item vectors
explore_items <- scale_converted %>% select(starts_with("explore")) %>% colnames()
expoit_items <- scale_converted %>% select(starts_with("exploit")) %>% colnames()
deimp_items <- scale_converted %>% select(starts_with("deimp")) %>% colnames()

all_items <- c(explore_items, expoit_items, deimp_items)

# calculate the polychoric correlation matrix
poly_results <- psych::polychoric(scale_converted[, all_items], na.rm = TRUE)
poly_matrix  <- poly_results$rho


# setup clean labels for the matrix
clean_labels <- c(
  c(paste("Explore", 1:7), paste("Explore def", 1:2)),
  c(paste("Exploit", 1:6), paste("Exploit def", 1:2)),
  c(paste("Deimplement", 1:4), paste("Deimplement def", 1:2))
)

rownames(poly_matrix) <- clean_labels
colnames(poly_matrix) <- clean_labels

# create the Corrplot
poly_matrix %>% ggcorrplot(type = "full",
                           outline.col = "white",
                           colors = c("#E46726", "white", "#6D9EC1"),
                           legend.title = "Polychoric Correlation",
                           tl.cex = 8) +
  # add boxes to show groups
  geom_rect(aes(xmin = 0.5, xmax = 9.5, ymin = 0.5, ymax = 9.5),
            fill = NA, color = "royalblue3", size = 0.5) +
  geom_rect(aes(xmin = 9.5, xmax = 17.5, ymin = 9.5, ymax = 17.5),
            fill = NA, color = "royalblue3", size = 0.5) +
  geom_rect(aes(xmin = 17.5, xmax = 23.5, ymin = 17.5, ymax = 23.5),
            fill = NA, color = "royalblue3", size = 0.5) + 
  labs(x = "", y = "", 
       caption = "Figure1: Polychoric correlation matrix for exploration, exploitation and deimplementaion") +
  theme_minimal() + 
  theme(
    plot.caption = element_text(hjust = 0.5, size = 9, colour = "darkgrey"),
    axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1, size = 7),
    axis.text.y = element_text(size = 7),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 7),
    )
```


# Create Final Dataset

- Drop variables that have no variability (have only one value)
- Calculate row-wise completion rate

```{r create final df}
final_data <- demog_final %>% 
  select(-c(language_oath_v2_ASL, language_oath_v2_LSQ, sign_languages)) %>% 
  right_join(scale_converted, by = "record_id") %>% 
  select(-c(missing_rate, prop_8)) %>% 
  # calculate overall completion rate
  mutate(completion_rate = round(rowMeans(!is.na(.))*100, 2))

dim(final_data)
```

## Check Completion Rate

```{r completion rate}
final_data %>% 
  summarise(
    avg = mean(completion_rate),
    min = min(completion_rate),
    max = max(completion_rate),
    prop_all_comp = mean(completion_rate == 100.00)*100
  ) %>% 
  kable(digits = 2,
        caption = "Row-wise Completion Rate Summary",
        col.names = c("Average", "Minimum", "Maximum", "All Completed (%)"),
        format = "html") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r save final data}
saveRDS(final_data, "Data/oath_data.rds")
```

