---
title: "Initial Data Inspection"
author: "Yunfei Yu, 1005976918"
date: "2026-02-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r libraries, message=FALSE}
library(readxl)
library(dplyr)
library(tidyverse)
library(kableExtra)
library(knitr)
library(skimr)
library(naniar)
library(VIM)
```

```{r import helper functions, message=FALSE}
source("helper_functions.R")
```

# 1 Outline

## Scales

7-point Likert Scale. 8 = Not Applicable.

1. Check distributions of scores for each item(including 8 and missing). 
2. Check missing values (percentage missing in each survey item, and missing patterns across all items and all scales).
  2.1 Explore demographic differences between the missing vs. non-missing groups.
  2.2 Explore demographic difference between the "actual missing (NA)" group vs. people who chose 8.
3. Compute mean scores and standard deviations for each item after removing NAs and 8.

## Demographics

1. Attach labels to the demographic dummy variables (i.e. "collapsing" multiple dummies into a categorical variable)
2. Check distributions
3. Check missing values and missing patterns
4. Compute summary statistics



```{r load data}
raw.data <- read_excel("Data/FINAL_ProjectOATH_DATA_2024-05-08_1639.xlsx")
# raw.data %>% skim()
```

# 2 Scales

This section checks the distributions and handles missing values for the survey scales, namely, the **Organizational Ambidexterity Survey**, **De-implementation Survey**, Organizational Capacity for Innovation Survey and Organizational Learning Survey.


## 2.1 Organizational Ambidexterity Survey - Exploration

### 2.1.1 Check distributions

```{r explore scale answer dist, fig.cap="Distribution of exploration survey scores"}
# plot distributions for each item in exploration survey
plt_scale_dist(raw.data, prefix = "explore")
```

### 2.1.2 Check missing and completion rates

```{r exploration scale missing}
# check missingness in exploration scale items
# get_missing_stats is a helper function from helper_functions.R
get_missing_stats(data = raw.data, prefix = "explore", caption = "Exploration scale completion rate")
```



## 2.2 Organizational Ambidexterity Survey - Exploitation

### 2.2.1 Check distributions

```{r exploit scale answer dist, fig.cap="Distribution of exploitation survey scores"}
# plot distributions for each item in exploitation survey
plt_scale_dist(raw.data, prefix = "exploit")
```


### 2.2.2 Check missing and completion rates

```{r exploitation scale missing}
# check missingness in exploitation scale items
get_missing_stats(data = raw.data, prefix = "exploit", caption = "Exploitation scale completion rate")
```


## 2.3 De-implementation Survey

### 2.3.1 Check distributions

```{r deimp scale answer dist, fig.cap="Distribution of deimplementation survey scores"}
# plot distributions for each item in deimp survey
plt_scale_dist(raw.data, prefix = "deimplement")
```



### 2.3.2 Check missing and completion rates

```{r deimp scale missing}
# check missingness in exploitation scale items
get_missing_stats(data = raw.data, prefix = "deimplement", caption = "De-implementation scale completion rate")
```


# 3 Demographics

## 3.1 Attaching labels to demographic variables

### Demographics - Site and Role

- Primary site `site_oath_v2`: categorical (factor)

- Role `role_oath_v2`: categorical (factor)

  - `role_oath_spec` if `role_oath_v2` = 7: text
  
- Number of years in current role `years_role_oath_v2`: categorical (factor)

- Number of years at THP `years_employed_thp_v2`: categorical (factor)

- Have academic appoint `academic_appt`: binary



# 4 Missing Values

## 4.1  Missingness Pattern across Exploration, Exploitation and De-Implementation Surveys

```{r scales upset plot, fig.height=12, out.width="100%"}
learning.scales <- raw.data %>% 
  select(starts_with(c("explore", "exploit", "deimplement")))

gg_miss_upset(learning.scales, nsets = ncol(learning.scales))
```

The Learning Ambidexterity scales and De-implementation scale appear to show highly systematic patterns of missingness. The most common pattern is skipping (missing) all questions in all three scales, followed by those who filled out only the exploration survey but missed the exploitation and de-implementation surveys entirely, and finally those who filled out the exploration and exploitation surveys, but not the de-implementation survey.
  
This suggests missingness is likely driven by an external factor (ex. specific groups of non-respondents, lost interest, or a specific survey is less relevant), rather than missing at random.











