---
title: "Initial Data Inspection"
author: "Yunfei Yu, 1005976918"
date: "2026-02-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r libraries, message=FALSE}
library(readxl)
library(dplyr)
library(tidyverse)
library(kableExtra)
library(knitr)
library(skimr)
library(naniar)
# library(VIM)
library(gtsummary)
library(gt)
library(forcats)
library(stringr)
```

```{r import helper functions, message=FALSE}
source("helper_functions.R")
```

# 1 Outline

## Scales

7-point Likert Scale. 8 = Not Applicable.

1. Check distributions of scores for each item (including 8 and missing). 

2. Compute mean scores and standard deviations for each item after removing NAs and 8.

## Demographics

1. Attach labels to the demographic variables
  a. For MC: convert multiple dummies to one factor variable
  b. For checkbox: explore ways for dimension reduction (e.g., combining groups)
2. Check distributions and summary statistics



```{r load data}
raw.data <- read_excel("Data/FINAL_ProjectOATH_DATA_2024-05-08_1639.xlsx")
# raw.data %>% skim()
```

```{r unique identifier check}
# check for any duplicates in record id
nrow(raw.data) == n_distinct(raw.data$record_id)
```


# 2 Scales

This section checks the distributions and handles missing values for the survey scales, namely, the **Organizational Ambidexterity Survey**, **De-implementation Survey**, Organizational Capacity for Innovation Survey and Organizational Learning Survey.


## 2.1 Organizational Ambidexterity Survey - Exploration

### 2.1.1 Check distributions

```{r explore scale answer dist, fig.cap="Distribution of exploration survey scores"}
# plot distributions for each item in exploration survey
plt_scale_dist(raw.data, prefix = "explore")
```

### 2.1.2 Check missing and completion rates

```{r exploration scale missing}
# check missingness in exploration scale items
# get_missing_stats is a helper function from helper_functions.R
get_missing_stats(data = raw.data, prefix = "explore", caption = "Exploration scale completion rate")
```



## 2.2 Organizational Ambidexterity Survey - Exploitation

### 2.2.1 Check distributions

```{r exploit scale answer dist, fig.cap="Distribution of exploitation survey scores"}
# plot distributions for each item in exploitation survey
plt_scale_dist(raw.data, prefix = "exploit")
```


### 2.2.2 Check missing and completion rates

```{r exploitation scale missing}
# check missingness in exploitation scale items
get_missing_stats(data = raw.data, prefix = "exploit", caption = "Exploitation scale completion rate")
```


## 2.3 De-implementation Survey

### 2.3.1 Check distributions

```{r deimp scale answer dist, fig.cap="Distribution of deimplementation survey scores"}
# plot distributions for each item in deimp survey
plt_scale_dist(raw.data, prefix = "deimplement")
```


### 2.3.2 Check missing and completion rates

```{r deimp scale missing}
# check missingness in exploitation scale items
get_missing_stats(data = raw.data, prefix = "deimplement", caption = "De-implementation scale completion rate")
```


# 3 Demographics

## 3.1 Site and Role

- Primary site `site_oath_v2`: categorical (factor)

- Role `role_oath_v2`: categorical (factor)

  - `role_oath_spec` if `role_oath_v2` = 7: text
  
- Number of years in current role `years_role_oath_v2`: categorical (factor)

- Number of years at THP `years_employed_thp_v2`: categorical (factor)

- Have academic appoint `academic_appt`: binary

```{r site and role labelling}
labelled.data <- raw.data %>% 
  # convert to factor and attach text labels for summary
  mutate(
    site_oath_v2 = factor(site_oath_v2,
                          levels = c(1, 2, 3, 4),
                          labels = c("Credit Valley Hospital", 
                                     "Missisauga Hospital", 
                                     "Queensway Health Centre", 
                                     "Other")
                          ),
    role_oath_v2 = factor(role_oath_v2,
                          levels = c(1, 2, 3, 4, 5, 6, 7),
                          labels = c("Administrative/Consultant",
                                     "Director",
                                     "Manager",
                                     "Professional Staff",
                                     "Senior Leadership",
                                     "Supervisor/Team Leader",
                                     "Other")
                          ),
    years_role_oath_v2 = factor(years_role_oath_v2,
                                levels = c(1, 2, 3, 4, 5),
                                labels = c("<1 year",
                                           "1-3 years",
                                           "3-6 years",
                                           "6-9 years",
                                           "10+ years")
                                ),
    years_employed_thp_v2 = factor(years_employed_thp_v2,
                                   levels = c(1, 2, 3, 4, 5),
                                   labels = c("<1 year",
                                              "1-3 years",
                                              "3-6 years",
                                              "6-9 years",
                                              "10+ years")
                                   ),
    academic_appt = factor(academic_appt,
                           levels = c(1, 0),
                           labels = c("Yes", "No"))
  )
```


```{r role self specify}
# check self-defined role 

# make sure this field is not empty only when role = 7/Other
# labelled.data %>% 
#   summarise(n = sum((role_oath_v2 != "Other") & (!is.na(role_oath_spec))))

get_category_summary(labelled.data %>% select(role_oath_spec),
                     caption = "Self-defined role")

# Create a new category for educator
# and merge "practice leader" into "Supervisor/Team Leader"
# drop role_oath_spec
labelled.data <- labelled.data %>% 
  mutate(role_oath_v2 = case_when(
    coalesce(str_detect(role_oath_spec, regex("educator", ignore_case = TRUE)), FALSE) ~ "Educator",
    coalesce(str_detect(role_oath_spec, regex("leader", ignore_case = TRUE)), FALSE) ~ "Supervisor/Team Leader",
    TRUE ~ role_oath_v2
  )) %>% 
  select(-role_oath_spec) 

# The remaining "Other" are those who didn't specify
```


```{r site and role summary}
# Missings are treated as an explicit category only for reporting
get_category_summary(labelled.data %>% 
                       select(site_oath_v2, role_oath_v2, years_role_oath_v2, years_employed_thp_v2, academic_appt),
                     caption = "Summary statistics (n (%)) for site and role variables",
                     id = "record_id")
```





## 3.2 Demographics

- Age `age_oath_v2`: categorical

- Gender `gender_oath_v2`: checkbox

  - Self-definition of gender `self_define_gender` only if `gender_oath_v2` = 5: text
  
- Sexual orientation `sexual_orient_oath_v2`: checkbox

  - Self definition of sexual orientation `self_define_sex1` only if `sexual_orient_oath_v2` = 7: text
  
- Disability `disability_oath_v2`: categorical

- Indigenous `indigenous_oath_v2`: categorical

- Visible minority / people of colour `poc_oath_v2`: categorical

- Language(s) `language_oath_v2`: checkbox (select all apply)

  - Specify `language_specify_oath_v2` only if `language_oath_v2` = 3 or 16: text
  
- Race `race_oath_v2`: checkbox (select all apply)

  - Specify `race_spec` only if `race_oath_v2` = 7: text

### 3.2.1 Factor variables: age, disability, indigenous, poc
  
```{r categorical demog labelling}
# attach labels to categorical columns
# i.e. age, disability, indigenous, vis min
labelled.data <- labelled.data %>% 
  mutate(
    age_oath_v2 = factor(age_oath_v2,
                         levels = c(1, 2, 3, 4, 5, 6),
                         labels = c("<18 years",
                                    "18-29 years",
                                    "30-54 years",
                                    "55-64 years",
                                    "65+ years",
                                    "Prefer not to answer")
                         ),
    disability_oath_v2 = factor(disability_oath_v2,
                                levels = c(1, 2, 3),
                                labels = c("Yes", "No", "Prefer not to answer")
                                ),
    indigenous_oath_v2 = factor(indigenous_oath_v2,
                                levels = c(1, 2, 3),
                                labels = c("Yes", "No", "Prefer not to answer")
                                ),
    poc_oath_v2 = factor(poc_oath_v2,
                         levels = c(1, 2, 3),
                         labels = c("Yes", "No", "Prefer not to answer")
                         )
  )
```


### 3.2.2 Checkbox (MC): gender and sexual orientation

```{r checkbox gender and sex}
# check if gender and sexual orientation is MC or choose all apply
labelled.data <- labelled.data %>% 
  mutate(gender_sum = rowSums(pick(starts_with("gender_"))),
         sex_sum = rowSums(pick(starts_with("sexual_")))) %>% 
  # treat as missing if sum = 0 (selected none)
  mutate(
    gender_sum = case_when(
      gender_sum == 0 ~ NA,
      TRUE ~ gender_sum
      ),
    sex_sum = case_when(
      sex_sum == 0 ~ NA,
      TRUE ~ sex_sum
    )
  )

# summary table
get_category_summary(labelled.data %>%
                       select(gender_sum, sex_sum),
                     caption = "Total score for gender and sexual orientation")


# if gender = 5, and if self-define is empty ~ other; if self-define not empty ~ self-defined text

```
  
Gender and sexual orientation are multiple choice in nature (rather than "select all that apply"), except for one person who chose more than one sexual orientation. Some participants did not select any categories for gender and sexual orientations.


```{r sexual orientation check}
# check the one person who choose more than 1 sexual orientation
labelled.data %>% 
  filter(sex_sum > 1)

# sex orient 3: straight/heterosexual
# sex orient 5: queer
```

```{r gender sex no selection check}
# check people who did not choose any category
# check row-wise completion rate
labelled.data %>% 
  filter(is.na(gender_sum)| is.na(sex_sum))
  
# same group of people
```


```{r check gender and sex self define}
get_category_summary(labelled.data %>%
                       select(self_define_gender, self_define_sex1), 
                     caption = "Self-defined gender and sexual orientation missing rate")
# No one self-defined gender or sexual orientation
```


```{r labelling gender and sex}
# convert checkbox gender and sexual orientation to factor
# if selecting more than 1 gender/sexual orientation -> treat as missing

labelled.data <- labelled.data %>% 
  mutate(
    gender_oath_v2 = case_when(
      gender_sum > 1 ~ NA,
      gender_oath_v2___1 == 1 ~ "Woman",
      gender_oath_v2___2 == 1 ~ "Man",
      gender_oath_v2___3 == 1 ~ "Trans",
      gender_oath_v2___4 == 1 ~ "Two Spirit",
      gender_oath_v2___5 == 1 ~ "Prefer to self-define",
      gender_oath_v2___6 == 1 ~ "Prefer not to answer",
      TRUE ~ NA
    ),
    sexual_orient_oath_v2 = case_when(
      sex_sum > 1 ~ NA, # who chose more than 1 sexual orientation
      sexual_orient_oath_v2___1 == 1 ~ "Gay",
      sexual_orient_oath_v2___2 == 1 ~ "Lesbian",
      sexual_orient_oath_v2___3 == 1 ~ "Straight/Heterosexual",
      sexual_orient_oath_v2___4 == 1 ~ "Bisexual",
      sexual_orient_oath_v2___5 == 1 ~ "Queer",
      sexual_orient_oath_v2___6 == 1 ~ "Two Spirit",
      sexual_orient_oath_v2___7 == 1 ~ "Prefer to self-define",
      sexual_orient_oath_v2___8 == 1 ~ "Prefer not to answer",
      TRUE ~ NA
    )
  ) %>% 
  # drop original cols
  select(-starts_with(c("gender_oath_v2___", "sexual_orient_oath_v2___"))) %>% 
  # drop sum cols and self spec cols
  select(-c(sex_sum, gender_sum, self_define_gender, self_define_sex1))
```


### 3.2.3 Checkbox (Select all that apply): language and race

```{r checkbox language and race}
# check number of language / race
labelled.data <- labelled.data %>% 
  mutate(num_language = rowSums(pick(starts_with("language_oath"))),
         num_race = rowSums(pick(starts_with("race_oath")))) %>% 
  mutate(
    num_language = case_when(
      num_language == 0 ~ NA,
      TRUE ~ num_language
      ),
    num_race = case_when(
      num_race == 0 ~ NA,
      TRUE ~ num_race
    )
  )

get_category_summary(labelled.data %>%
                       select(num_language, num_race), 
                     caption = "Number of language / race")
```


```{r check language and race self define}
# check self-defined language and race

# first convert to title case
labelled.data <- labelled.data %>% 
  mutate(language_specify_oath_v2 = as.factor(str_to_title(language_specify_oath_v2)),
         race_spec = as.factor(str_to_title(race_spec)))

get_category_summary(labelled.data %>%
                       select(language_specify_oath_v2, race_spec),
                     caption = "Self-defined language and race missing rate")
```

#### Language

```{r labelling language}
# for dimension reduction, keep the top 5 most spoken language and combine all others into "other"

# rename language cols
labelled.data <- labelled.data %>% 
  rename(
    language_oath_v2_English = language_oath_v2___1,
    language_oath_v2_French = language_oath_v2___2,
    language_oath_v2_Indigenous = language_oath_v2___3,
    language_oath_v2_Mandarin = language_oath_v2___4,
    language_oath_v2_Cantonese = language_oath_v2___5,
    language_oath_v2_Punjabi = language_oath_v2___6,
    language_oath_v2_Hindi = language_oath_v2___7,
    language_oath_v2_Urdu = language_oath_v2___8,
    language_oath_v2_Spanish = language_oath_v2___9,
    language_oath_v2_Portuguese = language_oath_v2___10,
    language_oath_v2_Tagalog = language_oath_v2___11,
    language_oath_v2_Vietnamese = language_oath_v2___12,
    language_oath_v2_Arabic = language_oath_v2___13,
    language_oath_v2_ASL = language_oath_v2___14,
    language_oath_v2_LSQ = language_oath_v2___15,
    language_oath_v2_Other = language_oath_v2___16,
    language_oath_v2_PNA = language_oath_v2___17,
  )

# check frequency
get_category_summary(labelled.data %>%
                       select(starts_with("language_oath")),
                     caption = "Frequency of language spoken")

```

```{r check language frequency}
# check frequency
labelled.data %>% 
  select(starts_with("language_oath_v2")) %>% 
  summarise(across(everything(), \(x) round(mean(x, na.rm = TRUE), 4)*100)) %>% 
  pivot_longer(everything(),
               names_to = "Language",
               values_to = "Percentage") %>% 
  mutate(Language = str_remove(Language, "language_oath_v2_")) %>% 
  filter(Language != "Other" & Language != "PNA") %>% # excluding other and PNA in the ranking
  arrange(desc(Percentage)) %>% 
  head(5) %>% 
  # create html table
  kable(digits = 2,
        col.names = c("Language", "Frequency (%)"),
        caption = "Top 5 most spoken language",
        format = "html") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r combine less spoken languages into other}
# For languages that are not among the top 5 most spoken,
# combine into the "Other" category

# The new "Other" category: 1 if any of the languages other than the top 5 and PNA is 1, 0 otherwise

pattern <- "English|French|Hindi|Cantonese|Punjabi|PNA"
languages.other <- labelled.data %>%
  select(starts_with("language_oath_v2") & !matches(pattern), language_specify_oath_v2) %>% 
  names()

labelled.data <- labelled.data %>% 
  mutate(language_oath_v2_AllOther = if_else(
    # any of the 'other' columns are 1 (ignoring NAs)
    if_any(all_of(languages.other), ~ .x == 1 & !is.na(.x)) | 
    # the specify column is not NA
    !is.na(language_specify_oath_v2), 
    1, 0, 
    missing = 0
  )) %>% 
  select(-languages.other)

# sanity check if language_oath_v2_AllOther is created correctly
# labelled.data %>%
#   select(languages.other, language_oath_v2_AllOther) %>%
#   mutate(language_specify_oath_v2 = if_else(is.na(language_specify_oath_v2), 0, 1)) %>% 
#   # calculate row sum
#   mutate(check = if_else(rowSums(across(languages.other)) == 0, 0, 1)) %>% 
#   summarise(n_incorrect = sum(language_oath_v2_AllOther != check))


```

```{r check language freq after combining}
# check frequency
labelled.data %>% 
  select(starts_with("language_oath_v2")) %>% 
  summarise(across(everything(), \(x) round(mean(x, na.rm = TRUE), 4)*100)) %>% 
  pivot_longer(everything(),
               names_to = "Language",
               values_to = "Percentage") %>% 
  mutate(Language = str_remove(Language, "language_oath_v2_")) %>% 
  arrange(desc(Percentage)) %>%
  # create html table
  kable(digits = 2,
        col.names = c("Language", "Frequency (%)"),
        caption = "Language frequency after combining categories",
        format = "html") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"))
```

#### Race

```{r labelling race}
# for dimension reduction, keep the top 5 most frequent race and combine all others into "other"

# rename language cols
labelled.data <- labelled.data %>% 
  rename(
    race_oath_v2_Asian = race_oath_v2___1,
    race_oath_v2_Black = race_oath_v2___2,
    race_oath_v2_LatinHispanic = race_oath_v2___3,
    race_oath_v2_MiddleEastern = race_oath_v2___4,
    race_oath_v2_Mixed = race_oath_v2___5,
    race_oath_v2_White = race_oath_v2___6,
    race_oath_v2_Other = race_oath_v2___7,
    race_oath_v2_PNA = race_oath_v2___8,
  )

# check frequency
get_category_summary(labelled.data %>%
                       select(starts_with("race_oath")),
                     caption = "Frequency of race")

```

```{r check race frequency}
# check frequency
labelled.data %>% 
  select(starts_with("race_oath_v2")) %>% 
  summarise(across(everything(), \(x) round(mean(x, na.rm = TRUE), 4)*100)) %>% 
  pivot_longer(everything(),
               names_to = "Race",
               values_to = "Percentage") %>% 
  mutate(Race = str_remove(Race, "race_oath_v2_")) %>% 
  # filter(Race != "Other" & Race != "PNA") %>% # excluding other and PNA in the ranking
  arrange(desc(Percentage)) %>% 
  # head(5) %>% 
  # create html table
  kable(digits = 2,
        col.names = c("Race", "Frequency (%)"),
        caption = "Race frequency",
        format = "html") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"))
```


```{r clean up specified race}
# labelled.data %>%
#   filter(!is.na(race_spec)) %>%
#   select(record_id, starts_with("race_"))

labelled.data <- labelled.data %>% 
  rename(race_oath_v2_BroadAsian_or_AsianHeritage = race_oath_v2_Asian) %>% 
  mutate(
    # combine South Asian, Indo-Caribbean and West Indian into the broader Asian / Asian heritage category
    # for aggregation purposes
    race_oath_v2_BroadAsian_or_AsianHeritage = case_when(
      coalesce(str_detect(race_spec, regex("Asian|Indian|Caribbean", ignore_case = TRUE)), FALSE) ~ 1,
      TRUE ~ race_oath_v2_BroadAsian_or_AsianHeritage
    ),
    # assuming Canadian and European are White
    race_oath_v2_White = case_when(
      coalesce(str_detect(race_spec, regex("Canadian|European", ignore_case = TRUE)), FALSE) ~ 1,
      TRUE ~ race_oath_v2_White
    ),
    # for these subgroups convert "Other" to 0
    race_oath_v2_Other = case_when(
      coalesce(str_detect(race_spec, regex("Asian|Indian|Caribbean", ignore_case = TRUE)), FALSE) ~ 0,
      coalesce(str_detect(race_spec, regex("Canadian|European", ignore_case = TRUE)), FALSE) ~ 0,
      TRUE ~ race_oath_v2_Other
    )
  ) %>% 
  select(-race_spec)
```


```{r check race frequency after merging}
# check frequency
labelled.data %>% 
  select(starts_with("race_oath_v2")) %>% 
  summarise(across(everything(), \(x) round(mean(x, na.rm = TRUE), 4)*100)) %>% 
  pivot_longer(everything(),
               names_to = "Race",
               values_to = "Percentage") %>% 
  mutate(Race = str_remove(Race, "race_oath_v2_")) %>% 
  # filter(Race != "Other" & Race != "PNA") %>% # excluding other and PNA in the ranking
  arrange(desc(Percentage)) %>% 
  # head(5) %>% 
  # create html table
  kable(digits = 2,
        col.names = c("Race", "Frequency (%)"),
        caption = "Race frequency",
        format = "html") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"))
```



```{r}
# glimpse(labelled.data)
```

```{r export data}
write_csv(labelled.data, "Data/OATH_DATA_Labelled.csv")
```







