---
title: "Initial Data Inspection"
author: "Yunfei Yu, 1005976918"
date: "2026-02-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r libraries, message=FALSE}
library(readxl)
library(dplyr)
library(tidyverse)
library(kableExtra)
library(knitr)
library(skimr)
library(naniar)
# library(VIM)
library(gtsummary)
library(gt)
library(forcats)
library(stringr)
```

```{r import helper functions, message=FALSE}
source("helper_functions.R")
```

# 1 Outline

## Scales

7-point Likert Scale. 8 = Not Applicable.

1. Check distributions of scores for each item (including 8 and missing). 
2. Check missing values (percentage missing in each survey item, and missing patterns across all items and all scales).
  2.1 Explore demographic differences between the missing vs. non-missing groups.
  2.2 Explore demographic difference between the "actual missing (NA)" group vs. people who chose 8.
3. Compute mean scores and standard deviations for each item after removing NAs and 8.

## Demographics

1. Attach labels to the demographic (checkbox) dummy variables (i.e. "collapsing" multiple dummies into a categorical variable)
2. Check distributions
3. Check missing values and missing patterns
4. Compute summary statistics


```{r load data}
raw.data <- read_excel("Data/FINAL_ProjectOATH_DATA_2024-05-08_1639.xlsx")
# raw.data %>% skim()
```

```{r unique identifier check}
# check for any duplicates in record id
nrow(raw.data) == n_distinct(raw.data$record_id)
```


# 2 Scales

This section checks the distributions and handles missing values for the survey scales, namely, the **Organizational Ambidexterity Survey**, **De-implementation Survey**, Organizational Capacity for Innovation Survey and Organizational Learning Survey.


## 2.1 Organizational Ambidexterity Survey - Exploration

### 2.1.1 Check distributions

```{r explore scale answer dist, fig.cap="Distribution of exploration survey scores"}
# plot distributions for each item in exploration survey
plt_scale_dist(raw.data, prefix = "explore")
```

### 2.1.2 Check missing and completion rates

```{r exploration scale missing}
# check missingness in exploration scale items
# get_missing_stats is a helper function from helper_functions.R
get_missing_stats(data = raw.data, prefix = "explore", caption = "Exploration scale completion rate")
```



## 2.2 Organizational Ambidexterity Survey - Exploitation

### 2.2.1 Check distributions

```{r exploit scale answer dist, fig.cap="Distribution of exploitation survey scores"}
# plot distributions for each item in exploitation survey
plt_scale_dist(raw.data, prefix = "exploit")
```


### 2.2.2 Check missing and completion rates

```{r exploitation scale missing}
# check missingness in exploitation scale items
get_missing_stats(data = raw.data, prefix = "exploit", caption = "Exploitation scale completion rate")
```


## 2.3 De-implementation Survey

### 2.3.1 Check distributions

```{r deimp scale answer dist, fig.cap="Distribution of deimplementation survey scores"}
# plot distributions for each item in deimp survey
plt_scale_dist(raw.data, prefix = "deimplement")
```


### 2.3.2 Check missing and completion rates

```{r deimp scale missing}
# check missingness in exploitation scale items
get_missing_stats(data = raw.data, prefix = "deimplement", caption = "De-implementation scale completion rate")
```


# 3 Demographics

## 3.1 Site and Role

- Primary site `site_oath_v2`: categorical (factor)

- Role `role_oath_v2`: categorical (factor)

  - `role_oath_spec` if `role_oath_v2` = 7: text
  
- Number of years in current role `years_role_oath_v2`: categorical (factor)

- Number of years at THP `years_employed_thp_v2`: categorical (factor)

- Have academic appoint `academic_appt`: binary

```{r site and role labelling}
labelled.data <- raw.data %>% 
  # convert to factor and attach text labels for summary
  mutate(
    site_oath_v2 = factor(site_oath_v2,
                          levels = c(1, 2, 3, 4),
                          labels = c("Credit Valley Hospital", 
                                     "Missisauga Hospital", 
                                     "Queensway Health Centre", 
                                     "Other")
                          ),
    role_oath_v2 = factor(role_oath_v2,
                          levels = c(1, 2, 3, 4, 5, 6, 7),
                          labels = c("Administrative/Consultant",
                                     "Director",
                                     "Manager",
                                     "Professional Staff",
                                     "Senior Leadership",
                                     "Supervisor/Team Leader",
                                     "Other")
                          ),
    years_role_oath_v2 = factor(years_role_oath_v2,
                                levels = c(1, 2, 3, 4, 5),
                                labels = c("<1 year",
                                           "1-3 years",
                                           "3-6 years",
                                           "6-9 years",
                                           "10+ years")
                                ),
    years_employed_thp_v2 = factor(years_employed_thp_v2,
                                   levels = c(1, 2, 3, 4, 5),
                                   labels = c("<1 year",
                                              "1-3 years",
                                              "3-6 years",
                                              "6-9 years",
                                              "10+ years")
                                   ),
    academic_appt = factor(academic_appt,
                           levels = c(1, 0),
                           labels = c("Yes", "No"))
  )
```


```{r site and role summary}
# Missings are treated as an explicit category only for reporting
get_category_summary(labelled.data %>% 
                       select(site_oath_v2, role_oath_v2, years_role_oath_v2, years_employed_thp_v2, academic_appt),
                     caption = "Summary statistics (n (%)) for site and role variables",
                     id = "record_id")
```


## 3.2 Demographics

- Age `age_oath_v2`: categorical

- Gender `gender_oath_v2`: checkbox

  - Self-definition of gender `self_define_gender` only if `gender_oath_v2` = 5: text
  
- Sexual orientation `sexual_orient_oath_v2`: checkbox

  - Self definition of sexual orientation `self_define_sex1` only if `sexual_orient_oath_v2` = 7: text
  
- Disability `disability_oath_v2`: categorical

- Indigenous `indigenous_oath_v2`: categorical

- Visible minority / people of colour `poc_oath_v2`: categorical

- Language(s) `language_oath_v2`: checkbox (select all apply)

  - Specify `language_specify_oath_v2` only if `language_oath_v2` = 3 or 16: text
  
- Race `race_oath_v2`: checkbox (select all apply)

  - Specify `race_spec` only if `race_oath_v2` = 7: text

### 3.2.1 Factor variables: age, disability, indigenous, poc
  
```{r categorical demog labelling}
# attach labels to categorical columns
# i.e. age, disability, indigenous, vis min
labelled.data <- labelled.data %>% 
  mutate(
    age_oath_v2 = factor(age_oath_v2,
                         levels = c(1, 2, 3, 4, 5, 6),
                         labels = c("<18 years",
                                    "18-29 years",
                                    "30-54 years",
                                    "55-64 years",
                                    "65+ years",
                                    "Prefer not to answer")
                         ),
    disability_oath_v2 = factor(disability_oath_v2,
                                levels = c(1, 2, 3),
                                labels = c("Yes", "No", "Prefer not to answer")
                                ),
    indigenous_oath_v2 = factor(indigenous_oath_v2,
                                levels = c(1, 2, 3),
                                labels = c("Yes", "No", "Prefer not to answer")
                                ),
    poc_oath_v2 = factor(poc_oath_v2,
                         levels = c(1, 2, 3),
                         labels = c("Yes", "No", "Prefer not to answer")
                         )
  )
```


### 3.2.2 Checkbox (MC): gender and sexual orientation

```{r checkbox gender and sex}
# check if gender and sexual orientation is MC or choose all apply
labelled.data <- labelled.data %>% 
  mutate(gender_sum = rowSums(pick(starts_with("gender_"))),
         sex_sum = rowSums(pick(starts_with("sexual_")))) %>% 
  # treat as missing is sum = 0 (selected none)
  mutate(
    gender_sum = case_when(
      gender_sum == 0 ~ NA,
      TRUE ~ gender_sum
      ),
    sex_sum = case_when(
      sex_sum == 0 ~ NA,
      TRUE ~ sex_sum
    )
  )

# summary table
get_category_summary(labelled.data %>%
                       select(gender_sum, sex_sum),
                     caption = "Total score for gender and sexual orientation")


# if gender = 5, and if self-define is empty ~ other; if self-define not empty ~ self-defined text

```
  
Gender and sexual orientation are multiple choice in nature (rather than "select all that apply"), except for one person who chose more than one sexual orientation. Some participants did not select any categories for gender and sexual orientations.


```{r sexual orientation check}
# check the one person who choose more than 1 sexual orientation
labelled.data %>% 
  filter(sex_sum > 1)

# sex orient 3: straight/heterosexual
# sex orient 5: queer
```

```{r gender sex no selection check}
# check people who did not choose any category
# check row-wise completion rate
labelled.data %>% 
  filter(is.na(gender_sum)| is.na(sex_sum))
  
# same group of people
```


```{r check gender and sex self define}
get_category_summary(labelled.data %>%
                       select(self_define_gender, self_define_sex1), 
                     caption = "Self-defined gender and sexual orientation missing rate")
# No one self-defined gender or sexual orientation
```


```{r labelling gender and sex}
# convert checkbox gender and sexual orientation to factor
# if selecting more than 1 gender/sexual orientation -> treat as missing

labelled.data <- labelled.data %>% 
  mutate(
    gender_oath_v2 = case_when(
      gender_sum > 1 ~ NA,
      gender_oath_v2___1 == 1 ~ "Woman",
      gender_oath_v2___2 == 1 ~ "Man",
      gender_oath_v2___3 == 1 ~ "Trans",
      gender_oath_v2___4 == 1 ~ "Two Spirit",
      gender_oath_v2___5 == 1 ~ "Prefer to self-define",
      gender_oath_v2___6 == 1 ~ "Prefer not to answer",
      TRUE ~ NA
    ),
    sexual_orient_oath_v2 = case_when(
      sex_sum > 1 ~ NA, # who chose more than 1 sexual orientation
      sexual_orient_oath_v2___1 == 1 ~ "Gay",
      sexual_orient_oath_v2___2 == 1 ~ "Lesbian",
      sexual_orient_oath_v2___3 == 1 ~ "Straight/Heterosexual",
      sexual_orient_oath_v2___4 == 1 ~ "Bisexual",
      sexual_orient_oath_v2___5 == 1 ~ "Queer",
      sexual_orient_oath_v2___6 == 1 ~ "Two Spirit",
      sexual_orient_oath_v2___7 == 1 ~ "Prefer to self-define",
      sexual_orient_oath_v2___8 == 1 ~ "Prefer not to answer",
      TRUE ~ NA
    )
  ) %>% 
  # drop original cols
  select(-starts_with(c("gender_oath_v2___", "sexual_orient_oath_v2___"))) %>% 
  # drop sum cols
  select(-c(sex_sum, gender_sum))
```


### 3.2.3 Checkbox (Select all that apply): language and race

```{r checkbox language and race}
# check number of language / race
labelled.data <- labelled.data %>% 
  mutate(num_language = rowSums(pick(starts_with("language_oath"))),
         num_race = rowSums(pick(starts_with("race_oath")))) %>% 
  mutate(
    num_language = case_when(
      num_language == 0 ~ NA,
      TRUE ~ num_language
      ),
    num_race = case_when(
      num_race == 0 ~ NA,
      TRUE ~ num_race
    )
  )

get_category_summary(labelled.data %>%
                       select(num_language, num_race), 
                     caption = "Number of language / race")
```


```{r check language and race self define}
# check self-defined language and race

# first convert to title case
labelled.data <- labelled.data %>% 
  mutate(language_specify_oath_v2 = as.factor(str_to_title(language_specify_oath_v2)),
         race_spec = as.factor(str_to_title(race_spec)))

get_category_summary(labelled.data %>%
                       select(language_specify_oath_v2, race_spec),
                     caption = "Self-defined language and race missing rate")
```


```{r labelling language}
# for dimension reduction, keep the top 5 most spoken language and combine all others into "other"

# check frequency
get_category_summary(labelled.data %>%
                       select(starts_with("language_oath")),
                     caption = "Frequency of language spoken")

```



# 4 Missing Values

## 4.1 Row-wise Missing

```{r}
# check missing percentage (completion rate) for each observation
```


## 4.2 Missingness Pattern across Exploration, Exploitation and De-Implementation Surveys

```{r scales upset plot, fig.height=12, out.width="100%"}
learning.scales <- labelled.data %>% 
  select(starts_with(c("explore", "exploit", "deimplement")))

gg_miss_upset(learning.scales, nsets = ncol(learning.scales))
```

The Learning Ambidexterity scales and De-implementation scale appear to show highly systematic patterns of missingness. The most common pattern is skipping (missing) all questions in all three scales, followed by those who filled out only the exploration survey but missed the exploitation and de-implementation surveys entirely, and finally those who filled out the exploration and exploitation surveys, but not the de-implementation survey.
  
This suggests missingness is likely driven by an external factor (ex. specific groups of non-respondents, lost interest, or a specific survey is less relevant), rather than missing at random.


## 4.3  Missingness Pattern for Demographics

```{r demographic site and role upset plot, fig.height=6, out.width="100%"}
demog_role <- labelled.data %>% 
  select(site_oath_v2, role_oath_v2, years_role_oath_v2, years_employed_thp_v2, academic_appt)
gg_miss_upset(demog_role, nsets = ncol(learning.scales))
```








