---
title: "Latent Profile Analysis"
author: "Yunfei Yu"
date: "2026-06-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, message=FALSE}
library(tidyverse)
library(tidyLPA)
library(ggplot2)
```

```{r import helper functions, message=FALSE}
source("helper_functions.R")
```

```{r load data}
oath_data <- readRDS("Data/oath_data_fa_final.rds")
lpa_data <- oath_data %>% 
  select(Explore, Exploit, Deimplement) %>% 
  drop_na()

dim(oath_data)
```


# Fit LPA

```{r fit lpa}
# Fit Model 1 (Equal variances)
models_1_fit <- lpa_data %>% estimate_profiles(n_profiles = 1:5, models = 1)

# Fit Model 2 (Varying variances)
models_2_fit <- lpa_data %>% estimate_profiles(n_profiles = 1:5, models = 2)

# Print the fit indices table
models_1_fit
models_2_fit

lpa_models <- models_2_fit
```

- Model 2 generally give lower AIC/BIC and higher Entropy than Model 1

- 4 profiles is the optimal in terms of AIC/BIC and Entropy, but risks of having a very small size for profile 4 (n_min = 0.05 (about 9 person))

- 3 profile model is the second optimal choice

# Comparison

```{r plot 3 profiles}
# profile_3 <- lpa_models[[3]]
# profile_4 <- lpa_models[[4]]
# 
# # Now plot them side-by-side
# plot_profiles(profile_3)
# plot_profiles(profile_4)
```

## Line plots

```{r cleaner plot}
# std mean score: std dev from the mean

# Plot the 3-profile solution
plot_lpa_profiles(lpa_models[[3]], plot_title = "LPA: 3-Profile Solution")

# Plot the 4-profile solution
plot_lpa_profiles(lpa_models[[4]], plot_title = "LPA: 4-Profile Solution")
```


```{r check profile size}
get_profile_sizes(lpa_models[[3]])
get_profile_sizes(lpa_models[[4]])
```


### 3-Profile Model

Pros: 
- Stable Group Sizes: more balanced distribution (46%, 43%, and 11%).

- Methodologically Sound: The smallest group has $N = 17$, maintaining adequate statistical power for downstream analysis.

- Clear Interpretation: Yields clean, non-crossing trajectories representing distinct, actionable strategies.

Cons: 

-Higher AIC/BIC and lower entropy than the 4-profile model.

### 4-Profile Model

Pros: 
- Best statistical fit, with the lowest AIC/BIC and highest Entropy.

Cons: 

- Small groups: Over-splits the 11% group into two tiny subgroups of $N = 7$ (4.5%) and $N = 10$ (6.5%). Groups this small lack statistical power for future hypothesis testing and are often artifacts of outlier noise.

- Confusion: Profiles cross on the plot, creating overly complex interaction narratives.


# LPA Results

(Based on 3-profile model)

- **Profile 1 - The Status-Quo** (N = 71, 46%): Slightly above average across all three dimensions. Highly consistent, average-level effort allocation across de-implementation, exploration and exploitation.

- **Profile 2 - De-implementation Drivers** (N = 17, 11%): Above average across all three dimensions. Scores exceptionally high on deimplementation, and also the highest on exploration and exploitation. 

- **Profile 3 - Restrained** (N = 65, 42%): Below average across all three dimensions. Scores exceptionally low on deimplementation. Allocating more effort towards exploitation.


# Store output

```{r}
profile_assignments <- lpa_models[[3]]$model$classification

regression_data <- oath_data %>%
  mutate(
    Profile = factor(profile_assignments)
  )

dim(regression_data)

saveRDS(regression_data, "Data/regression_data.rds")
```




