---
title: "Analysis"
author: "Yunfei Yu"
date: "2026-04-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Data

```{r libraries, message=FALSE}
library(tidyverse)
library(dplyr)
library(psych)
library(lavaan)
library(semPlot)
library(ggplot2)
library(tidyverse)
library(knitr)
library(kableExtra)
```


```{r load data}
oath_data <- readRDS("data/oath_data.rds")

# make sure 8=N/A is converted to missing
oath_data <- oath_data %>%
  mutate(across(starts_with(c("explore_", "exploit_", "deimplement_")), 
                ~na_if(., 8))) %>% 
  filter(if_any(starts_with(c("explore_", "exploit_", "deimplement_")), ~ !is.na(.)))


# oath_data <- oath_data %>% 
#   drop_na(all_of(
#     starts_with(c("explore", "exploit", "deimplement"))
#   ))

dim(oath_data)
```


# CFA

```{r fit cfa}
# extract relevant column names
explore_items <- oath_data %>% select(starts_with("explore")) %>% select(-contains("def")) %>% colnames()
exploit_items <- oath_data %>% select(starts_with("exploit")) %>% select(-contains("def")) %>% colnames()
deimp_items <- oath_data %>% select(starts_with("deimpl")) %>% select(-contains("def")) %>% colnames() 

# create model strs 
explore_items_str <- paste(explore_items, collapse = " + ")
exploit_items_str <- paste(exploit_items, collapse = " + ")
deimp_items_str <- paste(deimp_items, collapse = " + ")


# full model str

# Model 1:  3-factor model with exploration, exploitation and de-implementation
# as distinct dimensions
m3f <- paste0(
  'Exploration =~ ', explore_items_str, '\n',
  'Exploitation =~ ', exploit_items_str, '\n',
  'De_implement =~ ', deimp_items_str
)

# Model 2: 2-factor model with de-imp merged with exploitation 
m2f_exploit <- paste0(
  'Exploration =~ ', explore_items_str, '\n',
  'Merged_Factor =~ ', paste(c(exploit_items, deimp_items), collapse = " + ")
)

# Model 3: 2-factor model with de-imp merged with exploration
m2f_explore <- paste0(
  'Exploitation =~ ', exploit_items_str, '\n',
  'Merged_Factor =~ ', paste(c(explore_items, deimp_items), collapse = " + ")
)

# Model 4: 1-factor model with all items merged together
m1f <- paste0(
  'General_Factor =~ ', paste(c(explore_items, exploit_items, deimp_items), collapse = " + ")
)

# Fit the models and compare results
fit3f <- cfa(m3f, data = oath_data, ordered = TRUE, missing = "pairwise", estimator = "WLSMV")
fit2f_exploit <- cfa(m2f_exploit, data = oath_data, ordered = TRUE, missing = "pairwise", estimator = "WLSMV")
fit2f_explore <- cfa(m2f_explore, data = oath_data,ordered = TRUE, missing = "pairwise", estimator = "WLSMV")
fit1f <- cfa(m1f, data = oath_data, ordered = TRUE, missing = "pairwise", estimator = "WLSMV")

# Compare 3-factor vs. De-imp/Exploit merge
comp_exploit <- lavTestLRT(fit3f, fit2f_exploit)

# Compare 3-factor vs. De-imp/Explore merge
comp_explore <- lavTestLRT(fit3f, fit2f_explore)

# Compare 3-factor vs. 1-factor model
comp_1f <- lavTestLRT(fit3f, fit1f)
```


## Preliminary CFA Results

```{r 3f model summary}
summary(fit3f)
```

```{r cfa comparison results 1}
comp_exploit
```

```{r cfa comparison results 2}
comp_explore
```

- The 3-factor model with exploration, exploitation and de-implementation as three distinct pillars fits significantly better than either of the 2 factor models, and achieving lower AIC and BIC in both cases.

```{r}
comp_1f
```

## Loadings plot full

```{r loadings plot, fig.height=20}
# # Generate the plot
# png("Plots/CFA_Plot.png", width = 12, height = 10, units = "in", res = 300)
# 
# semPaths(fit3f, 
#          whatLabels = "std",       # Standardized loadings
#          layout = "tree2",         # Factors left, items right
#          rotation = 2,             # Horizontal layout
#          
#          # --- Spacing and Sizing ---
#          nCharNodes = 0,           # Don't truncate item names
#          sizeLat = 10,             # Size of factor ovals
#          sizeMan = 5,              # Smaller boxes for 23 items
#          edge.label.cex = 0.8,     # Smaller loading numbers to avoid overlap
#          
#          # --- Aesthetic ---
#          style = "lisrel",         # Professional black and white look
#          color = "white",
#          edge.color = "black",
#          shapeLat = "ellipse",
#          shapeMan = "rectangle",
#          
#          # --- Cleaning ---
#          residuals = FALSE,        # Hide error terms (arrows pointing to items)
#          intercepts = FALSE,       # Hide mean triangles
#          curvePivot = TRUE,        # Curve the correlations
#          
#          mar = c(3, 10, 3, 15))    # Large margins for long item names
# 
# dev.off() # Save the file    # Adjust margins to prevent clipping
```


## Refine CFA Model

- Drop items with loadings < 0.3

```{r small loading items}
# Get standardized parameters
params <- standardizedSolution(fit3f)

# find items where the loading is < 0.3
low_loading_items <- params %>%
  filter(op == "=~", est.std < 0.3) %>%
  select(rhs) %>%
  pull()

print(paste("Dropping items:", paste(low_loading_items, collapse = ", ")))
```


```{r refit cfa}
clean_explore <- setdiff(explore_items, low_loading_items)
clean_exploit <- setdiff(exploit_items, low_loading_items)
clean_deimp <- setdiff(deimp_items, low_loading_items)

# define model syntax
m3f_syntax <- paste0(
  'Exploration =~ ', paste(clean_explore, collapse = " + "), '\n',
  'Exploitation =~ ', paste(clean_exploit, collapse = " + "), '\n',
  'Deimplementation =~ ', paste(clean_deimp, collapse = " + ")
)

# 2-factor (merging exploit and deimp)
m2f_syntax_exploit <- paste0(
  'Exploration =~ ', paste(clean_explore, collapse = " + "), '\n',
  'Merged_Factor =~ ', paste(c(clean_exploit, clean_deimp), collapse = " + ")
)

# 2-factor (merging explore and deimp)
m2f_syntax_explore <- paste0(
  'Exploitation =~ ', paste(clean_exploit, collapse = " + "), '\n',
  'Merged_Factor =~ ', paste(c(clean_explore, clean_deimp), collapse = " + ")
)

# Fit the Clean Models
fit3f_clean <- cfa(m3f_syntax, data = oath_data, ordered = TRUE, missing = "pairwise", estimator = "WLSMV")
fit2f_clean_exploit <- cfa(m2f_syntax_exploit, data = oath_data, ordered = TRUE, missing = "pairwise", estimator = "WLSMV")
fit2f_clean_explore <- cfa(m2f_syntax_explore, data = oath_data, ordered = TRUE, missing = "pairwise", estimator = "WLSMV")
```

```{r rerun cfa comparison 1}
# 3f vs exploit-deimp merge
lavTestLRT(fit3f_clean, fit2f_clean_exploit)
```

```{r rerun cfa comparison 2}
# 3f vs explore-deimp merge
lavTestLRT(fit3f_clean, fit2f_clean_explore)
```


```{r final model fit}
# 1. Run the LRTs (Comparing 2-factor models against the 3-factor gold standard)
lrt_exploit <- lavTestLRT(fit2f_clean_exploit, fit3f_clean)
lrt_explore <- lavTestLRT(fit2f_clean_explore, fit3f_clean)

# 2. Extract and combine fit measures into a single data frame
fit_metrics <- c("df", "chisq", "cfi", "tli", "rmsea", "srmr")

summary_table <- bind_rows(
  as_tibble(as.list(fitMeasures(fit3f_clean, fit_metrics))) %>% 
    mutate(Model = "3-Factor"),
  as_tibble(as.list(fitMeasures(fit2f_clean_exploit, fit_metrics))) %>% 
    mutate(Model = "2-Factor<br>(Exploit-DI Merge)"),
  as_tibble(as.list(fitMeasures(fit2f_clean_explore, fit_metrics))) %>% 
    mutate(Model = "2-Factor<br>(Explore-DI Merge)")
)

# 3. Add the LRT p-values
# The p-value for the nested model is in the 2nd row of the lavTestLRT output
summary_table <- summary_table %>%
  mutate(
    `LRT p-val` = c(
      NA, 
      lrt_exploit$`Pr(>Chisq)`[2], 
      lrt_explore$`Pr(>Chisq)`[2]
    )
  ) %>%
  select(Model, df, chisq, `LRT p-val`, everything())

# 4. Final Formatting for your poster/report
summary_table %>%
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  mutate(`LRT p-val` = case_when(
    is.na(`LRT p-val`) ~ "",
    `LRT p-val` < 0.001 ~ "< .001",
    TRUE ~ as.character(`LRT p-val`)
  )) %>%
  kable(caption = "Table 1: CFA Model Fit Statistics and Nested Model Comparisons",
        digits = 3,
        col.names = c("Model", "Df", "Chisq", "LRT p-value", "CFI", "TLI", "RMSEA", "SRMR"),
        align = "c",
        format = "html",
        escape = FALSE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  add_header_above(c(" " = 1, "Model Fit" = 3, "Incremental Fit" = 2, "Residual Fit" = 2))
```



## Loadings plot reduced

```{r renaming cols for readability}
# clean up explore & exploit item names for plot readability
explore_names <- c("ER1", "ER2", "ER3", "ER5", "ER6") #  "ER_def1", "ER_def2"
exploit_names <- c("ET1", "ET2", "ET3", "ET5", "ET4", "ET6")
deimp_names <- c("DI1", "DI2", "DI3", "DI4") # "DI_def1", "DI_def2"

# update plot labels
# map original names to these new clean labels
explore_map <- setNames(explore_names, clean_explore)
exploit_map <- setNames(exploit_names, clean_exploit)
deimp_map <- setNames(deimp_names, clean_deimp)

# get the order of nodes from the lavaan object
actual_order <- row.names(inspect(fit3f_clean, "free")$lambda) 
latents <- slot(fit3f_clean, "Model")@dimNames[[1]][[2]]
all_factors <- c(actual_order, latents)

# create a lookup table for clean labels
lookup_table <- c(
  "Deimplementation" = "Deimplementation", 
  "Exploitation" = "Exploitation", 
  "Exploration" = "Exploration",
  explore_map, 
  exploit_map, 
  deimp_map
)

# create the final vector by looking up the labels in the correct order
final_node_labels <- ifelse(all_factors %in% names(lookup_table), 
                            lookup_table[all_factors], 
                            all_factors)
```


```{r reduced loadings plot}
groups_list <- list(Explore = clean_explore, 
                    Exploit = clean_exploit, 
                    Deimp = clean_deimp)

png("Plots/Clean_CFA.png", width = 10, height = 8, units = "in", res = 300)

semPaths(fit3f_clean,  
         nodeLabels = final_node_labels,
         whatLabels = "std",     # standardized loadings
         layout = "tree2",     
         style = "lisrel",
         nCharNodes = 0, 
         residuals = FALSE, 
         intercepts = FALSE,
         rotation = 2,
         thresholds = FALSE,
         
         # --- Sizing ---
         sizeLat = 14, 
         sizeMan = 6, 
         sizeLat2 = 5,
         sizeMan2 = 3,
         label.cex = 0.9,
         label.prop = 0.8,
         edge.label.cex = 0.6,
         edge.width = 0.5,

         # --- Grouping  ---
         edge.color = "#1A1953",
         groups = groups_list, 
         color = c("#ebf3f7", "#EFF5D0", "#ffe4c2"),
         legend = FALSE,

         # --- Spacing ---
         curvePivot = TRUE,      # Smooth curves for factor correlations
         
         # Left, Top, Bottom, Right
         # Setting a large left margin (20) to ensure items are not cut off.
         mar = c(3, 20, 3, 10)) 

dev.off() # Close the graphics device and save the file
```


## Factor scores 

```{r factor scores}
# Extract factor scores
factor_scores <- lavPredict(fit3f_clean)

# Convert to a dataframe and merge with oath_data
factor_scores_df <- as.data.frame(factor_scores)

# standardize
factor_scores_df <- factor_scores_df %>%
  mutate(across(c(Exploration, Exploitation, Deimplementation), 
                ~as.numeric(scale(.))))

# merge this back to your original data for the LPA
data_fa_scores <- cbind(oath_data, factor_scores_df)

saveRDS(data_fa_scores, "Data/oath_data_fa.rds")
```

```{r}

```

